The algebraic hierarchy in type theory using type classes

### 2010-06-14 jar.tex
Another snapshot. First 12 pages now more or less say what I want them to say.

@@

@@

\begin{document}
\title{The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\date{today}
\maketitle
\begin{abstract}
We present a new formalization of the algebraic hierarchy in Coq, exploiting its new type class mechanism to make practical a solution formerly thought infeasible. Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a bare minimum. To support mathematically sound abstract interfaces for $\N$, $\Z$, and $\Q$, our formalization includes portions of category theory and multisorted universal algebra.
Algebra flourishes by the interplay between syntax and semantics. The Prolog-like
abilities of type class unification allow us to conveniently define a quote function thus facilitating the use of reflective techniques.
\end{abstract}

@@

Several solutions have been proposed for the Coq theorem prover: dependent records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a. telescopes), packed classes~\cite{Packed}, and occasionally modules. We present a new approach based entirely on Coq's new type class mechanism, and show how its features together with a key design pattern let us effectively address the challenges mentioned above.

@@

\paragraph{Outline}
In section \ref{preliminaries} we briefly describe the Coq system and some of its facilities that are of particular interest to us (most notably its implementation of type classes). Then, in section \ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, as this is arguably the biggest design dimension when building interfaces for abstract structures. Having derived some principles concerning bundling, we then proceed to show (in section \ref{managing}) how type classes may be employed to adopt these principles without compromising ease of use. Next, in section \ref{hierarchy}, we introduce a number of additional conventions that let us complete a consistent idiom for interfaces to algebraic structures. With this in place, we continue with a discussion of

to the

@@

The ``native'' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where ``\lstinline|a ≡ b|'' is notation for \lstinline|eq T a b|~\footnote{We diverge from Coq traditional and reserve a single equals sign for setoid equality.}. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.

@@

In Haskell and Isabelle, type classes and their instances are second class, in that they are handled as specialized syntactic constructs whose semantics are given specifically by the type class apparatus. By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (``dictionaries''), instances are ordinary constants of these record types (registered as ``hints'' with the proof search machinery), class constraints are (fairly) ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.

Thus, type classes in Coq are realized by relatively minor syntactic aides that bring together existing facilities of the theory and the system into a coherent idiom, rather than by introduction of a new category of qualitatively different definitions with their own dedicated semantics.

% Often, we will want to use a \emph{different} equivalence relation when representing algebraic structures, for instance when representing integers by the hypothetical subtraction of a pair of natural numbers. A type and an equivalence relation over that type together form a \emph{setoid}.
% 
% An extremely important
% 
% 
% when doing algebra, however, one wants to use custom equivalence relations, for instance when
% 
% , for the purposes of doing algebra.
% 
% where term reduction is
% 
% the native
% 
% Leibniz equality 
% 
% (which in the Coq system amounts to
% 
% the ``native'' equality is term equality modulo reduction.
% 
% which amounts to 
% 
% normal forms
% 
% A trivial example being the method of representing integers by hypothetical subtractions of one natural from another.
% 
% 
% The Coq type theory lacks quotient
% types, as it would make type checking undecidable. % todo: reference
% Instead it supports setoids (aka Bishop sets), a pair of a type and an equivalence relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support for setoids has been added to the system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called \lstinline|Proper|.
% As we will see in Section~\ref{univ} below this functionality can also be used to work with
% quotient algebras.

@@

Algebraic structures are expressed in terms of a number of carriers, a number of operations and relations on these carriers, and a number of laws that the operations and relations satisfy. In a system like Coq, we have different options when it comes to representing the grouping of these components. On one end of the spectrum, we can simply define the (conjunction of) laws as a $n$-ary predicate over $n$ components, basically forgoing grouping altogether. For instance, for the mundane example of a reflexive relation, we could use:

\begin{lstlisting}
Definition reflexive {A} (R: relation A): Prop := Π a, R a a.
\end{lstlisting}
(The curly brackets used for \lstinline|A| mark it as an implicit argument.)

Higher structures, too, can all be expressed as predicates (expressing laws) over a number of carriers, relations, and operations. While optimally flexible in theory (todo: ref), in practice \emph{naive} adoption of this approach in Coq (i.e. without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest. And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or resort to awkward existentials. Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience. Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate convenient notations with them.

@@

In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\ref{corn} and Ssreflect\ref{ssreflect}), both use \emph{bundled} representations: records with one or more of the components as fields. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
Record ReflexiveRelation: Type :=
  { rr_carrier: Type
  ; rr_rel: relation rr_carrier
  ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it\footnote{While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``+'' and ``*'' notations to the record field projections for the addition and multiplication operations, respectively.}.

@@

Unfortunately, despite its virtues, the bundled representation introduces serious problems of its own, the most immediate and prominent one being a lack of support for \emph{sharing} components between structures.

In our example, lack of sharing support rears its head as soon as we try to define \lstinline|EquivalenceRelation| in terms of \lstinline|ReflexiveRelation| and its hypothetical siblings bundling symmetric and transitive relations. There, we would need some way to make sure that when we bundle the three constituents of types \lstinline|ReflexiveRelation|, \lstinline|SymmetricRelation|, and \lstinline|TransitiveRelation|, respectively together in a record, they all refer to the same carrier and relation. Positing additional equalities between the three bundled carriers and relations is neither easily accomplished (because one would need to work with heterogenous equality) nor would it permit natural use of the resulting structure (because one would constantly have to rewrite things back and forth).

Manifest records \ref{manifestrecords} have been proposed and implemented in the Matita system to address exactly this problem. We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution that avoids the problem altogether by employing type classes instead of bundled records.

If we revert back to the predicate formulation of reflexive (et.al) relations, we \emph{could} define \lstinline|EquivalenceRelation| in a bundled fashion without the need for equalities:
\begin{lstlisting}
Record EquivalenceRelation: Type :=
  { er_carrier: Type
  ; er_rel: relation er_carrier
  ; er_refl: ReflexiveRelation er_carrier er_rel
  ; er_sym: SymmetricRelation er_carrier er_rel
  ; er_trans: TransitiveRelation er_trans er_rel }.
\end{lstlisting}
However, the exact same argument inevitably leads to the conclusion that \lstinline|EquivalenceRelation|, too, should be a predicate. Indeed, it would be rather strange for the interface of equivalence relations to differ qualitatively from the interface of reflexive relations.

Another attempt to regain some grouping might be to bundle the carrier with the relation into a record. Unfortunately, this leads to the same problem of sharing, in different clothes: as soon as we try to define an algebraic structure with two reflexive relations on the same carrier, we need awkward hacks to establish equality between the carrier projections of two different (carrier, relation) bundles. To give a slightly less silly example of this phenomenon: bundling a monoid's constant and binary operation into a record means that when defining a hypothetical algebraic structure with two binary relations and a constant such that both binary relations form a monoid with the constant, one needs awkard hacks to establish equality of the two operation bundles' constants.

A second problem with bundling components is that as the bundled records are stacked to represent higher and higher structures, the projection paths for their components grow longer and longer, resulting in ever more unwieldy terms (although admittedly coercions and notations can make this less painful).

% hm, what to say about terms referring on proofs.

Thus, bundled representations come at a substantial cost in flexibility. Historically, this cost has been acceptable because (1) the unbundled alternative was such a pain, and (2) the standard algebraic hierarchy (up to, say, fields and modules) is not all that wild.

In the next section, we show that type-classification of structure predicates and their component parameters have the potential to remedy the problems traditionally associated with their use, giving us the best of both worlds.

For completeness, we mention an alternative record-based approach known as ``packed classes'', as used in the ssreflect library. In the packed classes idiom, each algebraic structure is represented by not one but a trinity of interdependent records along with packing/unpacking functions. These are then stacked in an intricate manner designed to reach a particular hybrid packing balance that, combined with the use of canonical structures and coercions, accomplishes some of the goals above.

We have found it hard to assess the general merits of this approach, partly because of its complexity, and partly because the only actual usage example (the hierarchy in the ssreflect library) is not only wedded to a very specialized view on equality and decidability that does not support setoid relations from the outset, but also diverges significantly from standard mathematical dichotomy, making it hard to recognize familiar structures and presentation. The reasons for this (as we understand them) are pragmatic and twofold.

First, the ssreflect library was originally developed strictly to support very practical work in a specialized domain of mathematics involving decidable and mostly finite structures. Second, at the time of ssreflect's initial development, support for setoid rewriting in Coq was not nearly as stable and refined as it is today, which made ``Leibniz equality at all costs'' a more reasonable strategy at the time than it might be considered today.

Work to extend the ssreflect framework with techniques to allow representation of a less restricted class of mathematical structures has only recently begun and is still ongoing, though there are currently no plans to support undecidable structures.

Because in our development we will be working with computable reals, equality on which can never be either Leibniz or decidable, we conclude that the ssreflect library is not usable for us.

@@

% We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
% \begin{lstlisting}

% Record Monoid (A: Type) (R: A → A → Prop)
%   (e: Equivalence R) : Type := ...
% \end{lstlisting}

% However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
% equivalence, we end up with two non-convertible ways to express that the relation (together with
% some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
% either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
% interchangable, and hope for automation, such as type class instance resolution, to use them
% interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
% essentially elevating them to `values', which goes against the principle of proof
% irrelevance.\marginpar{What if we add proof irrelevance?}
% BOTTOMLINE: Bundle as much as we can!
% 
% Neither of these options are very appealing. We really should just make |Monoid| take as
% parameters
% only the operations/relations. If we treat the monoid's unit and binary operation the same way,
% this
% means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
% how things ought to be.

@@

\section{Managing unbundled structure with type classes}

So far, we have only looked at the definitions of the structures themselves, and have not yet
considered their use when proving theorems about these structures.
We would like to approximate the common mathematical vernacular, where one
simply says:

for $x, y, z$ in a semigroup $G$, $x * y * z = z * y * x$, $\ldots$

Suppose we have defined \lstinline|SemiGroup| in line with the aforementioned principles:
\begin{lstlisting}
Record SemiGroup (A: Type) (eq: A $\to$ A $\to$ Prop)
 (op: A $\to$ A $\to$ A): Prop := ...
\end{lstlisting}

Then our starting point is the obvious but awful:
\begin{lstlisting}
Lemma perm A eq op (P: SemiGroup A eq op) x y z:
                   eq (op (op x y) z) (op (op z y) x).
\end{lstlisting}

As a first cleaning measure, we might hope to introduce some notations to let us use infix notation
for the semigroup operator (say, \lstinline|&|) and for equality. Unfortunately, at this point the names \lstinline|eq| and \lstinline|op| are just local names and there is no way to predeclare that they are to be associated with any notation. What we really need are canonical names for these notions, so that we may fashion notations for them.

This is a typical job for type classes, and they are used in this way in the \lstinline|Clean| standard
library. Similarly, Haskell uses type classes to overload notations.
We simply follow suit:
\begin{lstlisting}
  Class Equiv A := equiv: relation A.
  Class SemiGroupOp A := sg_op: A $\to$ A $\to$ A.

  Infix "=" := equiv: type_scope.
  Infix "&" := sg_op (at level 50, left associativity).
\end{lstlisting}

We choose to use \lstinline|&| here and reserve the notation \lstinline|*| for ring multiplication. We use = for setoid equality, since Leibniz equality seems less important for the formalization of
mathematics. More precisely, setoid equality is the basic concept. In some special cases, one can prove
that setoid equality coincides with Leibniz equality.
%Listings don't seem to work will with footnotes.
We would have like to use a haskell style definition
\begin{lstlisting}
Class Equiv A := (=): relation A.
\end{lstlisting}
Unfortunately, currently this is not supported in Coq.

These single-field type classes are given special treatment in that they are not translated into
records the way classes with any other number of fields are, but are instead turned into ordinary
definitions. The effect of this is that \lstinline|Equiv A| immediately reduces to \lstinline|relation A|, and that the equiv `projection' is the identity function whose argument is inferred using instance resolution. Thus, for a very modest cost in indirection, we gain the ability to use canonical names, and with it, notations:
\begin{lstlisting}
Record SemiGroup (A: Type) (eq: Equiv A)
                 (op: SemiGroupOp A): Prop := ...
Lemma perm A eq op (P: SemiGroup A eq op) x y z:
                 x & y & z = z & y & x.
\end{lstlisting}

We call these type classes \emph{operational type classes}. We note that operational type classes
allow us to avoid Coq's notation \lstinline|scope| mechanism.

Although the expression now \emph{looks} like it is bound to some
specific semigroup structure (which with traditional approaches would imply projections from bundles
and/or reference to proofs), it is really only referring near-directly to the actual operations
involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
may use in the proof. This lack of projections keeps our terms small and independent, and keeps
rewriting simple and sane; see Section~\ref{canonical}.

On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
context. The context for \lstinline|perm| looks like:
\begin{lstlisting}
  A: Type
  eq: Equiv A
  op: SemiGroupOp A
  P: SemiGroup A eq op
\end{lstlisting}

While we are not particularly worried about overly large contexts (we feel that this will most
likely not be problematic for any but the most complex formalizations), having to \emph{declare}
these entities (like \lstinline|eq| and \lstinline|op| for \lstinline|perm|) \emph{is} a bit of a chore. Fortunately, Coq provides a feature called implicit generalization, which is exactly what we need.
With it, we can write \lstinline|perm| as follows:
\begin{lstlisting}
  Lemma perm `(SemiGroup A) x y z: x & y & z = z & y & x.
\end{lstlisting}

Thus, we have reached the mathematical ideal we aimed for.

\section{Putting it all together}\label{classes}
Combining the ideas in the previous sections we end up with a formalization of algebraic
structures where we use the structure as (implicit) parameters, but
pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
\begin{lstlisting}
Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
                 {zero: RingZero A}{one: RingOne A}:Prop :=
  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
  ; semiring_plus_comm:> Commutative plus
  ; semiring_mult_comm:> Commutative mult
  ; semiring_distr:> Distribute mult plus
  ; mult_0_l: Π x, 0 * x = 0 }.
\end{lstlisting}

where e.g.\ \lstinline|RingPlus| is the operation class for the semigroup operator.
The notation \lstinline|:>| makes \lstinline|semiring_mult_monoid| a subclass: when asked for a monoid it suffices to
provide a ring.

Having argued that the \emph{all-structure-as-parameters} approach \emph{can} be made practical, we enumerate some of the benefits that make it worthwhile.

First, multiple inheritance becomes trivial: \lstinline|SemiRing| inherits two \lstinline|Monoid| structures on the same carrier and setoid relation, using ordinary named arguments to achieve ``manifest fields''; see section~\ref{manifest}.

Second, because our terms are small and independent and never refer to proofs, we are invulnerable to concerns about efficiency and ambiguity of projection paths that plague existing solutions, obviating the need for extensions like the proposed coercion pullbacks~\cite{Hints}.

Third, since our structural type classes are mere predicates, overlap between their instances is a non-issue. Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses: by simply declaring a \lstinline|SemiRing| class instance showing that a ring is a semiring, results about semirings immediately apply implicitly to any known ring, without us having to explicitly encode this relation in the hierarchy definition itself, and without needing any projection or translation of carriers or operations.

>>>>

\usepackage{graphicx}

@@

% preliminary choices:
%   don't discuss difference between maximally-inserted and non-maximally-inserted implicit arguments.

% todo:
  % what to say about terms referring on proofs?
  % cite that other type class based development in the context of the alternative bundling schemes.
  % maybe have a chapter called "interfaces are primary, not implementations" (also involve polynomials)
  % say this somewhere: \footnote{For an interesting (if fairly restrictive) alternative design in which unicity of term representations \emph{is} required for all objects, see the algebraic hierarchy in the ssreflect\cite{ssreflect} standard library.}

@@

\begin{document}
\title{The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\date{today}
\maketitle
\begin{abstract}
We present a new formalization of the algebraic hierarchy in Coq, exploiting its new type class mechanism to make practical a solution formerly thought infeasible. Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a bare minimum. To support mathematically sound abstract interfaces for $\N$, $\Z$, and $\Q$, our formalization includes portions of category theory and multisorted universal algebra.
Algebra flourishes by the interplay between syntax and semantics. The Prolog-like
abilities of type class unification allow us to conveniently define a quote function thus facilitating the use of reflective techniques.
  % todo: handle.
\end{abstract}

@@

Several solutions have been proposed for the Coq theorem prover: dependent records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a. telescopes), packed classes~\cite{Packed}, and occasionally modules. We present a new approach based entirely on Coq's new type class mechanism, and show how it can be used to make fully ``unbundled'' predicate-representations of algebraic structures practical to work with.

@@

\paragraph{Outline}
In section \ref{preliminaries} we briefly describe the Coq system and some of its facilities that are of particular interest to us (most notably its implementation of type classes). Then, in section \ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building 
interfaces for abstract structures. In section \ref{predicateclasses} we show how type classes can make the use of ``unbundled'' purely predicate based interfaces for abstract structures practical. Next, in section \ref{hierarchy}, we discuss our algebraic hierarchy implemented using such predicate classes.

 %Having derived some principles concerning bundling, we then proceed to show (in section \ref{managing}) how type classes may be employed to adopt these principles without compromising ease of use. Next, in section \ref{hierarchy}, we introduce a number of additional conventions that let us complete a consistent idiom for interfaces to algebraic structures. With this in place, we continue with a discussion of

%to the

@@

The ``native'' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where ``\lstinline|a ≡ b|'' is notation for \lstinline|eq T a b|~\footnote{We diverge from Coq traditional and reserve ``\lstinline|=|'' for setoid equality, as this is the equality we will be working with most of the time.}. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.

@@

In Haskell and Isabelle, type classes and their instances are second class, in that they are handled as specialized syntactic constructs whose semantics are given specifically by the type class apparatus. By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (``dictionaries''), instances are ordinary constants of these record types (registered as ``hints'' with the proof search machinery), class constraints are ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.

Thus, type classes in Coq are realized by relatively minor syntactic aids that bring together existing facilities of the theory and the system into a coherent idiom, rather than by introduction of a new category of qualitatively different definitions with their own dedicated semantics.

% The Coq type theory lacks quotient
% types, as it would make type checking undecidable. % todo: reference
% Instead it supports setoids (aka Bishop sets), a pair of a type and an equivalence relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support for setoids has been added to the system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called \lstinline|Proper|.
% As we will see in Section~\ref{univ} below this functionality can also be used to work with
% quotient algebras.

@@

Algebraic structures are expressed in terms of a number of carriers, a number of operations and relations on these carriers, and a number of laws that the operations and relations satisfy. In a system like Coq, we have different options when it comes to representing the grouping of these components. On one end of the spectrum, we can simply define the (conjunction of) laws as an $n$-ary predicate over $n$ components, forgoing explicit grouping altogether. For instance, for the mundane example of a reflexive relation, we could use:

\begin{lstlisting}
  Definition reflexive {A: Type} (R: relation A): Prop := Π a, R a a.
\end{lstlisting}
(The curly brackets used for \lstinline|A| mark it as an implicit argument.)

Higher structures, too, can all be expressed as predicates (expressing laws) over a number of carriers, relations, and operations. While optimally flexible in principle, in practice \emph{naive} adoption of this approach (i.e. without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest. And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time. Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience. Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.

@@

In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\ref{corn} and Ssreflect\ref{ssreflect}), both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it\footnote{While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``$+$'' and ``$*$'' notations to the record field projections for the addition and multiplication operations, respectively.}.

@@

Unfortunately, despite its apparent virtues, the bundled representation introduces serious problems of its own, the most immediate and prominent one being a lack of support for \emph{sharing} components between structures, which is needed to cope with overlapping multiple inheritance.

In our example, lack of sharing support rears its head as soon as we try to define \lstinline|EquivalenceRelation| in terms of \lstinline|ReflexiveRelation| and its hypothetical siblings bundling symmetric and transitive relations. There, we would need some way to make sure that when we ``inherit'' \lstinline|ReflexiveRelation|, \lstinline|SymmetricRelation|, and \lstinline|TransitiveRelation| by adding them as fields in our bundled record, they all refer to the same carrier and relation. Adding additional fields stating equalities between the three bundled carriers and relations is neither easily accomplished (because one would need to work with heterogenous equality) nor would it permit natural use of the resulting structure (because one would constantly have to rewrite things back and forth).

Manifest fields \ref{manifestrecords} have been proposed and implemented in the Matita system to address exactly this problem. We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution that avoids the problem altogether by using predicate-like type classes in place of bundled records.

If we revert back to the predicate formulation of reflexive (et. al) relations, we \emph{could} still define \lstinline|EquivalenceRelation| in a bundled fashion without the need for equalities:
\begin{lstlisting}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: relation er_carrier
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{lstlisting}
However, the same argument as before leads to the conclusion that \lstinline|EquivalenceRelation|, too, should be a predicate. Indeed, it would be rather strange for the interface of equivalence relations to differ qualitatively from the interface of reflexive relations.

Another attempt to recover some grouping might be to bundle the carrier with the relation into a (lawless) record, but this too hinders sharing: as soon as we try to define an algebraic structure with two reflexive relations on the same carrier, we need awkward hacks to establish equality between the carrier projections of two different (carrier, relation) bundles.

Even bundling just the operations of an algebraic structure together in a record (with the carrier as a parameter) does not work, as it again leads to the same problem when, for example, one attempts to define a hypothetical algebraic structure with two binary relations and a constant such that both binary relations form a monoid with the constant.

A second problem with bundling is that as the bundled records are stacked to represent higher and higher structures, the projection paths for their components grow longer and longer, resulting in ever more unwieldy terms (coercions and notations can make this less painful). Further, if one tries to implement some semblance of sharing in a bundled representation, these projection paths additionally become non-canonical, and still more extensions have been proposed to address this symptom (e.g. coercion pullbacks~\cite{Hints}).

Thus, bundled representations come at a substantial cost in flexibility. Historically, using bundled representations has nevertheless been an acceptable trade-off, because (1) the unbundled alternative was such a pain, and (2) the standard algebraic hierarchy (up to, say, fields and modules) is not all that wild.

In the next section, we show that type-classification of structure predicates and their component parameters has the potential to remedy the problems associated with the naive unbundled predicate approach, giving us the best of both worlds.

%For completeness, we mention an alternative record-based approach known as ``packed classes'', as used in the ssreflect library. In the packed classes idiom, each algebraic structure is represented by not one but a trinity of interdependent records along with packing/unpacking functions. These are then stacked in an intricate manner designed to reach a particular hybrid packing balance that, combined with the use of canonical structures and coercions, accomplishes some of the goals above.

%We have found it hard to assess the general merits of this approach, partly because of its complexity, and partly because the only actual usage example (the hierarchy in the ssreflect library) is not only wedded to a very specialized view on equality and decidability that does not support setoid relations from the outset, but also diverges significantly from standard mathematical dichotomy, making it hard to recognize familiar structures and presentation. The reasons for this (as we understand them) are pragmatic and twofold.

%First, the ssreflect library was originally developed strictly to support very practical work in a specialized domain of mathematics involving decidable and mostly finite structures. Second, at the time of ssreflect's initial development, support for setoid rewriting in Coq was not nearly as stable and refined as it is today, which made ``Leibniz equality at all costs'' a more reasonable strategy at the time than it might be considered today.

%Work to extend the ssreflect framework with techniques to allow representation of a less restricted class of mathematical structures has only recently begun and is still ongoing, though there are currently no plans to support undecidable structures.

%Because in our development we will be working with computable reals, equality on which can never be either Leibniz or decidable, we conclude that the ssreflect library is not usable for us.

@@

The observant reader may wonder whether it might be beneficial to go one step further and unbundle proof of laws and inherited substructures as well. This is not the case, because there is no point in sharing them. After all, by (weak) proof irrelevance, the ``value'' of such proofs can be of no consequence anyway. Indeed, parameterizing on proofs would be actively harmful because instantiations differing only in the proof argument would express the same thing yet be nonconvertible, requiring awkward conversions and hindering automation.

@@

\section{Predicate classes}

We will tackle the problems associated with the structure predicates approach one by one, starting with those encountered during theorem \emph{application}. Suppose we have defined \lstinline|SemiGroup| as a structure predicate as follows \footnote{Note that defining \lstinline|SemiGroup| as a record instead of a straight conjunction does not make it any less of a predicate. The record form is simply more convenient in that it immediately gives us named projections for laws and substructures.}:
\begin{lstlisting}
  Record SemiGroup (G: Type) (e: relation G) (op: G → G → G): Prop :=
    { sg_setoid: Equivalence e
    ; sg_ass: Associative op
    ; sg_proper: Proper (e ==> e ==> e) op }.
\end{lstlisting}
Then by (1) making \lstinline|SemiGroup| a \emph{class} (by replacing the \lstinline|Record| keyword with the \lstinline|Class| keyword), (2) marking its proofs as \emph{instances} (again by replacing a single keyword), and (3) marking the \lstinline|SemiGroup| parameter of semigroup theorems as implicit (by using curly instead of round brackets), we no longer have to pass \lstinline|SemiGroup| proofs around manually ourselves, letting instance resolution do it for us instead. Because instance resolution is part of the unifier, this also works when the statement of the theorem we wish to apply only mentions some of the components (which admittedly doesn't make much sense for semigroups).

Next, we turn to problems concerning theorem \emph{declaration}. As a reference point, our ideal will be the common mathematical vernacular, where one simply says:
\begin{quote}
Theorem L: For $x, y, z$ in a semigroup $G$, $x * y * z = z * y * x$.
\end{quote}
Without further support from the system, this would have to be written as
\begin{lstlisting}
  Theorem L G e op {P: SemiGroup G e op}:
    Π x y z, e (op (op x y) z) (op (op z y) x).
\end{lstlisting}
(The curly brackets mark \lstinline|P| as an implicit argument.) Because \lstinline|e| and \lstinline{op} are freshly introduced local names, we cannot bind notations to them prior to this theorem. Hence, if we want notations, what we really need are canonical names for these components. This is easily accomplished with single-field type classes containing one component each, which we will call \emph{operational type classes}\footnote{These single-field type classes are used in the same way in the \lstinline|Clean| standard library \cite{something}.}:
\begin{lstlisting}
  Class Equiv A := equiv: relation A.
  Class SemiGroupOp A := sg_op: A $\to$ A $\to$ A.

  Infix "=" := equiv: type_scope.
  Infix "&" := sg_op (at level 50, left associativity).
\end{lstlisting}
We use \lstinline|&| here and reserve the notation \lstinline|*| for (semi)ring multiplication.

As an aside, we note that the distinction between the class field name and the infix operator notation bound to it is really just a mildly awkward Coq artifact. In Haskell, where operators can themselves be used as names, there would be no need to have the \lstinline|equiv| and \lstinline|sg_op| names in addition to the operator ``names''.

If we now retype \lstinline|SemiGroup| as:
\begin{lstlisting}
  Π (G: Type) (e: Equiv G) (op: SemiGroupOp G), Prop
\end{lstlisting}
then we can declare the theorem with:
\begin{lstlisting}
  Theorem L G e op {P: SemiGroup G e op}: Π x y z, x & y & z = z & y & x.
\end{lstlisting}
This works because instance resolution, invoked by the use of \lstinline|=| and \lstinline|&|, will find \lstinline|e| and \lstinline|op|, respectively. Hence, the above is really

\begin{lstlisting}
  Theorem L G e op {P: SemiGroup G e op}: Π x y z,
    equiv e (sg_op op (sg_op op x y) z) (sg_op op (sg_op op x y) z).
\end{lstlisting}
where e and the op's are filled in by instance resolution.

At this point, a legitimate worry might be that the \lstinline|Equiv|/\lstinline|SemiGroup| classes and their \lstinline|equiv|/\lstinline|sg_op| projections imply constant construction and deconstruction of records, harming the simplicity and flexibility of the predicate approach that we are trying so hard to preserve. No such construction and destruction takes place, however, because type classes with only a single field are not desugared into an actual record with field projections the way classes with any other number of fields are. Instead, both class itself and its field projection are defined as the identity function with a fancy type. Thus, the introduction of these canonical names is essentially free; the structure predicate's new type reduces straight back to what it was before.

A remaining eyesore in the theorem declaration is the enumeration of \lstinline|e| and \lstinline|op|. To remove these, we use a new parameter declaration feature called \emph{implicit generalization}, introduced in Coq specifically to support type classes. Using implicit generalization, we can write:

\begin{lstlisting}
  Lemma L `{SemiGroup G}: Π x y z: G, x & y & z = z & y & x.
\end{lstlisting}
% TODO: describe implicit generalization in two words
% TODO: at some point down the line in our coq-inadequacies section, describe the implicit generalization improvements we need.
Thus, we have reached the mathematical ideal we aimed for.

% We note that operational type classes allow us to avoid Coq's notation \lstinline|scope| mechanism.
% hm, i don't think this is true. we don't need scopes for different representations' use of operators because we properly recognize their commonality, but that can be done without operational type classes as well.

% TODO: get this in:
% Although the expression now \emph{looks} like it is bound to some
% specific semigroup structure (which with traditional approaches would imply projections from bundles
% and/or reference to proofs), it is really only referring near-directly to the actual operations
% involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
% may use in the proof. This lack of projections keeps our terms small and independent, and keeps
% rewriting simple and sane; see Section~\ref{canonical}.

One aspect of the predicate approach we have not mentioned thus far is that in proofs parameterized by abstract structures, all components become hypotheses in the context. For the example theorem above, the context looks like:
\begin{lstlisting}
  G: Type
  e: Equiv G
  op: SemiGroupOp G
  P: SemiGroup G e op
\end{lstlisting}
We are not particularly worried about overly large contexts, especially because most of the ``extra'' hypotheses we have compared to bundled approaches are declarations of relations, operators, and constants, which are all in some sense ``inert'' with respect to proof search. Hence, we do not foresee problems with large contexts for any but the most complex formalizations.

\section{The algebraic hierarchy}\label{classes}
